Spherical harmonic analysis of faceted spheroids identifies shaping strategies and standardisation at Qianshangying (North China)

### 2025-08-22 analysis/paper/paper.qmd
edited to the end, let's see what I missed

For all spheroids and subspheroids, we conducted shaping efficiency analysis. \@fig-spearman-correlation-matrix shows significant positive correlations among the three variables representing reduction intensity: scar number, removal ratio, and SDI. Among them, scar counts show significant positive correlations with both removal ratio (ρ = `r cor_scar_removal`, p = $`r p_scar_removal`$) and SDI (ρ = `r cor_scar_sdi`, p = $`r p_scar_sdi`$). Removal ratio and SDI exhibit a stronger and more highly significant correlation (ρ = `r cor_removal_sdi`, p = $`r p_removal_sdi`$). These two variables incorporate surface area in their calculation, making them more effective in capturing the density and intensity of reduction, and therefore more reliable indicators of the shaping process in spheroids.

Among the shape descriptors of spheroids, several significant correlations were observed. Mean edge angle was positively correlated with sphericity (ρ = `r cor_angle_sph`, p = $`r p_angle_sph`$) and negatively correlated with Spherical Harmonics Power (SHE) (ρ = `r cor_angle_she`, p = $`r p_angle_she`$), indicating that greater edge angles are associated with more spherical and regular forms. As noted in previous studies, mean edge angle is a key variable in capturing the morphology of spheroids [@tittonSubspheroidsLithicAssemblage2020]. This correlation between edge angle and sphericity may help explain why spheroids were often interpreted as exhausted cores in earlier research [@sahnouni1997]. Indeed, spheroids must be significantly reduced, as achieving a near-perfect spherical shape typically results in increased edge angles, making further flake removal difficult. However, the reverse is not necessarily true, as not all exhausted cores evolve into spheroids.

@@

However, no significant correlations were observed between reduction variables and sphericity or curvature. This implies that increased reduction does not necessarily produce smoother surfaces or perfect spherical surfaces. Such patterns likely reflect the technological nature of the sample: the analyzed spheroids and sub-spheroids are "faceted" products of flake removal. Flaking increases surface complexity by adding edges and angular intersections, which may elevate curvature while also expanding surface area, factors that can distort sphericity values. For instance, a smooth, naturally rounded cobble might exhibit lower curvature and a sphericity closer to 1 than a heavily flaked spheroid, even if both appear similarly rounded overall. Therefore, the negative correlation between reduction intensity and SHE, but not sphericity, suggests that shaping efforts prioritized achieving symmetrical, coherent forms rather than polished, smooth surfaces.

@@

These findings also contribute to the long-standing debate over the nature of spheroids—whether they should be interpreted as byproducts of flake production, or as targeted end products of shaping strategies. In the case of the Qianshangying facted spheroids, the evidence strongly supports the latter. Faceted spheroids recovered from other Early Pleistocene sites in Barranco Leon and ‘Ubeidiya have also been reported to show evidence consistent with shaping strategy [@Muller2023; @tittonSubspheroidsLithicAssemblage2020]. These intentionally shaped spheroids appear contemporaneous with, or even predate, the emergence of Acheulean technology. It is important to note, however, that many spheroids from East Africa may not fall into this category. As noted by @mora2005, these quartz-based spheroids may have been produced through percussive techniques. They are often covered with battering marks, exhibit smoother surfaces, and lack ridges, features that suggest they may be better classified as bolas within a typological framework [@leakey1971]. It may be more appropriate to describe these artifacts in technological terms, rather than relying on morphology-based typologies that can lead to confusion. For instance, referring to them as battering spheroids could help distinguish them from faceted spheroids.

Another issue we aim to address is the relationship between polyhedrons, spheroids, and bolas. While no bolas were identified at Qianshangying, our quantitative analyses clearly differentiate polyhedrons from spheroids. These two artifact types differ in size, morphology, and production sequence. Polyhedrons at Qianshangying represent heavily reduced cores, whose edge angle increase results from extensive flake removal. Their primary purpose lies in producing flakes, placing them within a flaking strategy. In contrast, spheroids were intentionally shaped as end products, reflecting a shaping strategy. These two technological approaches of flaking and shaping strategies coexisted at Qianshangying, indicating a diversity of reduction goals.

@@

In East Asia, spheroids constitute a technological category with broad temporal and spatial distribution [@Yi2012Spheroids], as shown in @fig-Spheroids-distribution-China (see the supplementary materials for more information about these sites in table). In addition to the one spheroid collected near the Gongwangling site in Shanxi [@Dai1973Lantian], the earliest dated spheroids originate from Zhoukoudian Locality 1 (ZKD Loc.1) [@NIGPAS1985PekingMan], attributed to the early Middle Pleistocene, and not far from Qianshangying. In terms of geographic range, the southernmost and westernmost occurrences of spheroids are found at the Mujiaqiao site (MJQ) in Yunnan [@Wei1984], while the northernmost spheroids come from the Jinsitai site (JST) in Inner Mongolia [@Wang2010Jinstai]. The widespread occurrence of spheroids across time and space in East Asia may indicate their role as a generalized or multipurpose tool within Pleistocene contexts, capable of serving diverse functions in varying ecological and cultural settings.

Over 1522 spheroids have been identified across more than 23 archaeological sites or localities in East Asia However, many of these findings derive from surface surveys rather than systematic excavations, and the recovered quantities are often limited. Thus, the number of spheroids per site varies considerably: 19 sites have yielded fewer than ten specimens each. Qianshangying is typical with 'r n_spheroids' spheroids. By contrast, three sites—Xujiayao (XJY), Dingcun (DC), and JST—have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; @SXA2014Dingcun]. Notably, the XJY alone has yielded over 1,000 specimens [@Wang2016Houjiayao].

@@

@@

<!-- BM: for JAS I think we want the novelty of the methods to be the leading idea in the conclusion -->

Quantitative approaches to studying spheroids remain limited [@Muller2023; @deweyer2017; @tittonSubspheroidsLithicAssemblage2020]. We propose a methodological framework here represented by spherical harmonics to quantitatively assess morphological variation among flaked pieces, allowing for the distinction between cores and spheroids and evaluating the degree of morphological standardization in spheroids. This approach has proven effective and reliable. Past approaches to analyzing the morphology of flaked pieces have been limited, with geometric morphometrics measures being virtually the only method used for a long time [@hallinan2025; @cao2025; @lycett2013]. However, this method has significant constraints, including the requirement for homologous landmarks, which limits its applicability to artifacts like cores in shape variety and spheroids. Spherical harmonics offer a viable alternative, showing strong potential for precision and applicability. We can reasonably expect broader applications of this method across a range of archaeological artifacts in the future.

This study provides one of the earliest systematic analyses focused on spheroids from East Asia. By applying a range of quantitative methods, we conducted an integrated morphological and technological study of the Qianshangying spheroids. Our results demonstrate that these spheroids were produced through a programmed reduction strategy aimed at achieving standardized spherical forms, primarily using hard-hammer direct percussion. The evidence of shaping strategies on Qianshangying spheroids confirms that they were intentionally made tools with conceptual template rather than exhausted cores. This distinguishes them from polyhedrons or multi-facial cores in their terminal reduction stages. Traditionally, lithic technology in northern China prior to the Late Pleistocene has been considered stagnant and generally classified as simple core-and-flake industries. However, the Qianshangying assemblage from \~429ka clearly reveals a higher level of technological complexity

This paper also presents one of the first systematic overviews of the spatial and temporal distribution and technological variability of spheroids within China. We found that Chinese spheroids display considerable diversity in raw materials, production techniques, and morphology. Some assemblages contain spheroids that may be comparable to those from Qianshangying and tend to appear within Acheulean or shaping-oriented techno-complexes, often dated later than Qianshangying This highlights the particular significance of the Qianshangying spheroids. However, the current dataset remains limited, and many sites lack detailed morphological and technological studies of spheroids, hampering broader inferences about hominin technological behaviors. We therefore emphasize the need for site-specific analysis of spheroids in future research.

Given the limitations of traditional typology in capturing the functional, morphological, and technological variability of spheroids across sites, we argue that multiple method approaches are essential. From the perspective of typology, spheroids have often been dismissed as poor marker of hominin behavior due to their wide spatial and temporal distribution, combined with low-cost technologies, more refined studies increasingly reveal their diversity and potential as markers of homonin technological behavior [@tittonSubspheroidsLithicAssemblage2020; @vaquero2018]. For example, distinctions between flaking-based and battering-based spheroids, along with patterns of raw material selection, production goals, and associations with specific subsistence strategies or environmental adaptations, may prove meaningful [@Cabanes2024]. Beyond the methods of morphological and technological analysis used here, future studies incorporating functional analysis and experimental replication will further enhance our understanding.

>>>>

For all spheroids and subspheroids, we conducted shaping efficiency analysis. @fig-spearman-correlation-matrix shows significant positive correlations among the three variables representing reduction intensity: scar number, removal ratio, and SDI. Among them, scar counts show significant positive correlations with both removal ratio (ρ = `r cor_scar_removal`, p = $`r p_scar_removal`$) and SDI (ρ = `r cor_scar_sdi`, p = $`r p_scar_sdi`$). Removal ratio and SDI exhibit a stronger and more highly significant correlation (ρ = `r cor_removal_sdi`, p = $`r p_removal_sdi`$). These two variables incorporate surface area in their calculation, making them more effective in capturing the density and intensity of reduction, and therefore more reliable indicators of the shaping process in spheroids.

Among the shape descriptors of spheroids, we observed several significant correlations. Mean edge angle was positively correlated with sphericity (ρ = `r cor_angle_sph`, p = $`r p_angle_sph`$) and negatively correlated with Spherical Harmonics Power (SHE) (ρ = `r cor_angle_she`, p = $`r p_angle_she`$), indicating that greater edge angles are associated with more spherical and regular forms. As noted in previous studies, mean edge angle is a key variable in capturing the morphology of spheroids [@tittonSubspheroidsLithicAssemblage2020]. This correlation between edge angle and sphericity may help explain why spheroids were often interpreted as exhausted cores in earlier research [@sahnouni1997]. Indeed, spheroids must be significantly reduced, as achieving a near-perfect spherical shape typically results in increased edge angles, making further flake removal difficult. However, the reverse is not necessarily true, as not all exhausted cores evolve into spheroids.

@@

However, no significant correlations were observed between reduction variables and sphericity or curvature. This implies that increased reduction does not necessarily produce smoother surfaces or perfect spherical surfaces. Such patterns likely reflect the technological nature of the sample: the analyzed spheroids and sub-spheroids are "faceted" products of flake removal. Flaking increases surface complexity by adding edges and angular intersections, which may elevate curvature while also expanding surface area, factors that can distort sphericity values. For instance, a smooth, naturally rounded cobble might exhibit lower curvature and a sphericity closer to 1 than a heavily flaked spheroid, even if both appear similarly rounded overall. Therefore, the negative correlation between reduction intensity and SHE, but not sphericity, suggests that shaping efforts prioritized achieving symmetrical, coherent forms rather than naturally polished, smooth surfaces.

@@

These findings also contribute to the long-standing debate over the nature of spheroids--whether they should be interpreted as byproducts of flake production, or as targeted end products of shaping strategies. In the case of the Qianshangying facetted spheroids, the evidence strongly supports the latter. Faceted spheroids recovered from other Early Pleistocene sites in Barranco Leon and ‘Ubeidiya have also been reported to show evidence consistent with shaping strategy [@Muller2023; @tittonSubspheroidsLithicAssemblage2020]. These intentionally shaped spheroids appear contemporaneous with, or even predate, the emergence of Acheulean technology. It is important to note, however, that many spheroids from East Africa may not fall into this category. As noted by @mora2005, these quartz-based spheroids may have been produced through percussive techniques. They are often covered with battering marks, exhibit smoother surfaces, and lack ridges, features that suggest they may be better classified as bolas within a typological framework [@leakey1971]. It may be more appropriate to describe these artifacts in technological terms, rather than relying on morphology-based typologies that can lead to confusion. For instance, referring to them as battering spheroids could help distinguish them from faceted spheroids.

Another issue we aimed to address is the relationship between polyhedrons, spheroids, and bolas. While no bolas were identified at Qianshangying, our quantitative analyses clearly differentiated polyhedrons from spheroids. These two artifact types differ in size, morphology, and production sequence. Polyhedrons at Qianshangying represent heavily reduced cores, whose edge angle increase results from extensive flake removal. Their primary purpose lies in producing flakes, placing them within a flaking strategy. In contrast, spheroids were intentionally shaped as end products, reflecting a shaping strategy. These two technological approaches of flaking and shaping strategies coexisted at Qianshangying, indicating a diversity of reduction goals.

@@

In East Asia, spheroids constitute a technological category with broad temporal and spatial distribution [@Yi2012Spheroids], as shown in @fig-Spheroids-distribution-China (see the supplementary materials for a table of more information about these sites). In addition to the one spheroid collected near the Gongwangling site in Shanxi [@Dai1973Lantian], the earliest dated spheroids originate from Zhoukoudian Locality 1 (ZKD Loc.1) [@NIGPAS1985PekingMan], attributed to the early Middle Pleistocene, and not far from Qianshangying. In terms of geographic range, the southernmost and westernmost occurrences of spheroids are found at the Mujiaqiao site (MJQ) in Yunnan [@Wei1984], while the northernmost spheroids come from the Jinsitai site (JST) in Inner Mongolia [@Wang2010Jinstai]. The widespread occurrence of spheroids across time and space in East Asia may indicate their role as a generalized or multipurpose tool within Pleistocene contexts, capable of serving diverse functions in varying ecological and cultural settings.

Over 1522 spheroids have been identified across more than 23 archaeological sites or localities in East Asia However, many of these findings derive from surface surveys rather than systematic excavations, and the recovered quantities are often limited. Thus, the number of spheroids per site varies considerably: 19 sites have yielded fewer than ten specimens each. Qianshangying is typical with 'r n_spheroids' spheroids. By contrast, three sites--Xujiayao (XJY), Dingcun (DC), and JST—have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; @SXA2014Dingcun]. Notably, XJY alone has yielded over 1,000 specimens [@Wang2016Houjiayao].

@@

## Spheriods in a global cultural-chronological context

@@

Quantitative approaches to studying spheroids remain limited [@Muller2023; @deweyer2017; @tittonSubspheroidsLithicAssemblage2020]. We propose a complete, efficient, reproducible and transparent methodological framework here based on spherical harmonics to quantitatively assess morphological variation among flaked pieces, allowing for the distinction between cores and spheroids and evaluating the degree of morphological standardization in spheroids. Past approaches to analyzing the morphology of flaked pieces have been limited, with geometric morphometrics being the last major method innovation [@hallinan2025; @cao2025; @lycett2013]. However, this method has significant constraints, including the requirement for homologous landmarks, which limits its applicability to artifacts like cores of various shapes and spheroids. Spherical harmonics offer a viable alternative, showing strong potential for precision and intpretatbility. We can reasonably expect broader applications of this method across a range of archaeological artifacts in the future.

This study provides one of the first systematic analyses focused on spheroids from East Asia. By applying a range of quantitative methods, we conducted an integrated morphological and technological study of the Qianshangying spheroids. Our results demonstrate that these spheroids were produced through a programmed reduction strategy aimed at achieving standardized spherical forms, primarily using hard-hammer direct percussion. The evidence of shaping strategies on Qianshangying spheroids confirms that they were intentionally made tools with conceptual template, rather than exhausted cores. This distinguishes them from polyhedrons or multi-facial cores in their terminal reduction stages. Traditionally, lithic technology in northern China prior to the Late Pleistocene has been considered stagnant and generally classified as simple core-and-flake industries. However, the Qianshangying assemblage from \~429ka clearly reveals a higher level of technological complexity

This paper also presents one of the first systematic overviews of the spatial and temporal distribution and technological variability of spheroids within China. We found that Chinese spheroids display considerable diversity in raw materials, production techniques, and morphology. Some assemblages contain spheroids that may be comparable to those from Qianshangying and tend to appear within Acheulean or shaping-oriented techno-complexes, often dated later than Qianshangying. This highlights the particular significance of the Qianshangying spheroids. However, the current dataset remains limited, and many sites lack detailed morphological and technological studies of spheroids, hampering broader inferences about hominin technological behaviors. We therefore emphasize the need for site-specific analysis of spheroids in future research.

Given the limitations of traditional typology in capturing the functional, morphological, and technological variability of spheroids across sites, we argue that multiple method approaches are essential. From the perspective of typology, spheroids have often been dismissed as poor marker of hominin behavior due to their wide spatial and temporal distribution. However, more recent studies increasingly reveal their diversity and potential as markers of homonin technological behavior [@tittonSubspheroidsLithicAssemblage2020; @vaquero2018]. For example, distinctions between flaking-based and battering-based spheroids, along with patterns of raw material selection, production goals, and associations with specific subsistence strategies or environmental adaptations, may prove meaningful [@Cabanes2024]. Beyond the methods of morphological and technological analysis used here, future studies incorporating functional analysis and experimental replication will further enhance our understanding.

# CRediT authorship contribution statement

Yi Zhe: Writing – review & editing, Writing – original draft, Visualization, Resources, Project administration, Methodology, Investigation, Funding acquisition, Formal analysis, Data curation, Conceptualization. Ben Marwick: Writing – original draft, review & editing, Supervision, Project administration, Methodology, Conceptualization. Li Hao: Writing – review & editing, Conceptualization, Funding acquisition. Shuwen Pei: Writing – review & editing, Conceptualization, Funding acquisition.

# Data availability statement

The data that support the findings of this study are openly available at https://doi.org/xxxx/xxx

# Funding

This work was financially supported through the ...

# Declaration of competing interest

The authors declare that they have no known competing financial interests or personal relationships that could have appeared to influence the work reported in this paper.

# Acknowledgements

Thanks to Christine Harper for kindly sharing her expertise on SPHARM analyses.
